A Tale of Two Dictionary Learning Problems

### 2020-08-13 manuscripts/thesis/abstract.tex
redid abstract

\begin{abstract}
Learning optimal dictionaries for sparse coding has exposed characteristic sparse features of many natural signals. However, universal guarantees of the uniqueness and stability of such features in the presence of noise are lacking. This work presents very general conditions guaranteeing when dictionaries yielding the sparsest encodings of a dataset are unique and stable with respect to measurement or modeling error. The stability constants are explicit and computable; as such, there is an effective procedure sufficient to affirm if a proposed solution to the dictionary learning problem is unique within bounds commensurate with the noise. Beyond the extension of existing results to the noisy regime, a theory of combinatorial designs for sparse supports is introduced to demonstrate that some or all generating dictionary elements are recoverable from noisy data even if the dictionary fails to satisfy the spark condition, its size is overestimated, or only a polynomial number of distinct supports appear in the encoded data. Importantly, the guarantees derived here assume no constraints on the recovered dictionary beyond a natural upper bound on its size. The work closes with a lengthy discussion of the results and some remaining open challenges seeding future research directions.
\end{abstract}

>>>>

\begin{abstract}
Learning optimal dictionaries for sparse representation modeling has led to the discovery of characteristic sparse features in many natural signals. 
However, universal guarantees of the uniqueness and stability of such features in the presence of noise are lacking. 
This work presents very general conditions guaranteeing when dictionaries yielding the sparsest encodings of a dataset are unique and stable with respect to measurement or modeling error. 
The stability constants are explicit and computable; as such, there is an effective procedure sufficient to affirm if a proposed solution to the dictionary learning problem is unique within bounds commensurate with the noise.

Two formulations of the dictionary learning problem are considered. The first seeks a dictionary admitting a sparse representation of bounded support size for each point in a dataset. The second seeks a dictionary which minimizes the average induced support size over the dataset. In the first case, beyond the original extension of existing guarantees to the noisy regime, a theory of combinatorial designs for sparse supports is introduced to demonstrate that in almost all cases some or all dictionary elements are recoverable up to an error commensurate with the noise even if the dictionary fails to satisfy the spark condition, its size is overestimated, or data are distributed over only a polynomial number of subspaces spanned by the dictionary. These guarantees are then extended to the second case, becoming the first such guarantees in both in the noiseless and noisy regimes, by demonstrating that the second problem reduces to an instance of the first, given sufficient data. Importantly, in both cases the guarantees apply without imposing any assumptions at all on learned dictionaries beyond a natural upper bound on their size.

This work serves to justify, in principle, the application of dictionary learning to discover latent sparse structure in data; though much work remains to be done deriving practical criteria for use in applications. 
%The work closes with some open questions and directions for future research, seeded in part by the results of more practically-minded simulations. 
\end{abstract}

### 2020-08-13 manuscripts/thesis/discussion.tex
redid abstract

A commonly applied workaround to the intractability of Prob.~\ref{SLCopt} (see \cite{tillmann2015computational}) is to relax the sparsity measuring norm $\|\mathbf{x}\|_0$ to be $\|\mathbf{x}\|_1$, thereby transforming the inference of sparse $\mathbf{\overline x}_i$ for a fixed $\mathbf{B}$ into a convex optimization solvable by a linear program. Recalling the definition of $\mathbf{z}_i$ in \eqref{LinearModel}, the approach is rather:

\begin{problem}\label{ConvexifiedOptimizationProblem}
Find a matrix $\mathbf{B}$ and vectors \mbox{$\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$} solving:
\begin{equation}\label{l1minsum}
\min \sum_{i = 1}^N \|\mathbf{\overline x}_{i}\|_1 \ \
\text{subject to} \ \ \|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta, \ \text{for all $i$}.
\end{equation}
\end{problem}

A major advance in compressive sensing was the discovery that, when $\mathbf{B}$ is held fixed, solutions to \eqref{l1minsum} are in fact solutions to \eqref{minsum} provided $L_{2k}(\mathbf{B})$ is large enough \cite{eldar2012compressed}. Interestingly, the current work provides conditions on the generating dictionary $\mathbf{A}$ and $k$-sparse codes $\mathbf{x}_i$ under which \emph{all} matrices $\mathbf{B}$ (of bounded column-norm) that solve Prob.~\ref{OptimizationProblem} have $L_{2k}(\mathbf{B})$ bounded from below; specifically, $L_{2k}(\mathbf{B}) \geq \left(L_{2k}(\mathbf{A}) - C_1\varepsilon \right) / \|\mathbf{D}\|_1$ in the case where $\mathbf{A}$ satisfies \eqref{SparkCondition}. Thus, there is some noise bound inside of which all solutions to Prob.~\ref{OptimizationProblem} are solutions to the convexified problem as well. It is an open question as to whether practical constraints would exclude any remaining solutions to the convefixied problem that don't also solve Prob.~\ref{OptimizationProblem}.

As an intuitive example to kickstart these efforts, consider the simple noiseless ($\eta = 0$) case for $k=1$ where the matrix $\mathbf{A}$ has linearly independent unit-norm columns and $\mathbf{x}_i = c\mathbf{e}_i$ for some $c \geq 1$ for all $i \in [m]$. Suppose that for some $\mathbf{B}$ also having linearly independent unit-norm columns there are vectors $\mathbf{\overline x}_i$ satisfying $\|\mathbf{\overline x}_i\|_1 \leq c$ and $\mathbf{B \overline x}_i = \mathbf{Ax}_i$ for all $i \in [m]$. Intuitively, the only vectors in the convex polygon $\{ \mathbf{B \overline x}: \|\mathbf{\overline x}\|_1 \leq c\}$ for which $\|\mathbf{B \overline x}_i\|_2 = \|\mathbf{Ax}_i\|_2 = \|c\mathbf{Ae}_i\|_2 = c$ are its corners $\mathbf{B}(\pm c\mathbf{e}_j)$ for $j \in [m]$, so the $\mathbf{\overline x}_i$ must also be 1-sparse and we can apply Thm.~\ref{DeterministicUniquenessTheorem}.

>>>>

A commonly applied workaround to the intractability (see \cite{tillmann2015computational}) of Prob.~\ref{SLCopt} is to replace the norm $\|\mathbf{x}\|_0$ in \eqref{minsum} with its best convex approximation, $\|\mathbf{x}\|_1$, thereby transforming the inference of sparsest $\mathbf{\overline x}_i$ for fixed $\mathbf{B}$ into a convex optimization problem solvable by a linear program. Recalling the linear model \eqref{LinearModel} of data $\mathbf{z}_i$, the approach is to solve the following problem instead of attempting to solve Prob~\ref{SLCopt} directly.:

\begin{problem}\label{ConvexifiedOptimizationProblem}
Find a matrix $\mathbf{B}$ and vectors \mbox{$\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$} solving:
\begin{equation}\label{l1minsum}
\min \sum_{i = 1}^N \|\mathbf{\overline x}_{i}\|_1 \ \
\text{subject to} \ \ \|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta, \ \text{for all $i$},
\end{equation}
\end{problem}

A major advance in compressive sensing was the discovery that, when the matrix $\mathbf{B}$ is held fixed, the vectors $\mathbf{\overline x}_i$ that solve \eqref{l1minsum} also in fact minimize \eqref{minsum} provided $L_{2k}(\mathbf{B})$ is large enough \cite{eldar2012compressed}. Interestingly, the current work provides conditions on the generating dictionary $\mathbf{A}$ and $k$-sparse codes $\mathbf{x}_i$ under which \emph{all} matrices $\mathbf{B}$ solving Prob.~\ref{OptimizationProblem} have $L_{2k}(\mathbf{B})$ bounded from below; specifically, $L_{2k}(\mathbf{B}) \geq \left(L_{2k}(\mathbf{A}) - 2\eta C_1 \right) / \|\mathbf{D}\|_1$ in the case where $\mathbf{A}$ satisfies the spark condition \eqref{SparkCondition}. Thus, for suitable $\mathbf{A}$ there is some noise bound inside of which all solutions to Prob.~\ref{OptimizationProblem} are solutions to Prob.~\ref{ConvexifiedOptimizationProblem} as well. It is an open question as to whether there exist practical constraints which reject solutions to Prob.~\ref{ConvexifiedOptimizationProblem} that don't also solve Prob.~\ref{OptimizationProblem}.

We can already see this may be possible by examining the simple case $k=1$ without noise, i.e. $\eta = 0$.
%Consider the dataset generated as in \eqref{LinearModel} in the noiseless case $\eta = 0$, i.e.:
%\begin{align*}
%\mathbf{z}_i = \mathbf{A}\mathbf{x}_i,\ \ \  \text{$i=1,\ldots,N$}.
%\end{align*}
%with the additional constraint that $\|\mathbf{Ae}_i\|_2 = 1$ for all $i \in [m]$. 
%\begin{conjecture}
%Fix a matrix $\mathbf{A}$ with $\|\mathbf{A}\|_2=1$ and vectors $\mathbf{x}_i$ satisfying also the assumptions of Thm.~\ref{SLCopt}. Every solution to Prob.~\ref{ConvexifiedOptimizationProblem} satisfies BLAH and BLAH.
%\end{conjecture}

\begin{proposition}
If $\mathbf{A}$ has $L_2(\mathbf{A}) > 0$ and unit length-columns, and $\mathbf{x}_i = c_i\mathbf{e}_i$ with $c_i \neq 0$ for $i = 1, \ldots, m$, then every solution to Prob.~\ref{ConvexifiedOptimizationProblem} with matrices $\mathbf{B}$ having $m$ unit-length columns satisfies $\mathbf{A} = \mathbf{BP}$ and $\mathbf{x}_i = \mathbf{P}^\top \mathbf{\overline x}_i$ for some $m \times m$ permutation matrix $\mathbf{P}$.
\end{proposition}

\begin{proof}
We will show that the $\mathbf{\overline x}_i$ must in fact be $1$-sparse and then apply Thm.~\ref{DeterministicUniquenessTheorem}. Fix $i \in [m]$. Writing $\mathbf{\overline x}_i = \sum_{j=1}^m \overline c_{j} \mathbf{e}_j$, we have:
\begin{align}\label{gtc}
c_i = \|c_i\mathbf{Ae}_i\|_2 = \|\mathbf{B \overline x}_i\|_2 = \|\sum_{j=1}^m \overline c_{j} \mathbf{Be}_j\|_2 \leq \sum_{j=1}^m |\overline c_{j}| \|\mathbf{Be}_j\|_2 = \|\mathbf{\overline x}_i\|_1
\end{align}
So $\|\mathbf{\overline x}_i\|_1 \geq c_i$ for all $i \in [m]$. Therefore $\sum_{i=1}^m \|\mathbf{\overline x}_i\|_1 \geq \sum_i c_i$. But since $\mathbf{A}$ and the $\mathbf{x}_i$ satisfy the constraints of the minimization problem, we must also have $\sum_{i=1}^m \|\mathbf{\overline x}_i\|_1 \leq \sum_{i=1}^m \|\mathbf{x}_i\|_1 = \sum_i c_i$. Thus $\sum_{i=1}^m \|\mathbf{\overline x}_i\|_1 = \sum_i c_i$. Since again $\|\mathbf{\overline x}_i\|_1 \geq c_i$ for all $i \in [m]$, we must have $\| \mathbf{\overline x}_i\|_1 = c_i$ for all $i \in [m]$.

Revisiting \eqref{gtc}, we therefore have $c_i = \|\mathbf{B\overline x}_i\|_2 \leq \|\mathbf{\overline x}_i\|_1 = c_i$ for all $i \in [m]$, with equality holding only when $\overline c^{(i)}_j \mathbf{Be}_j$ are colinear. This would be the case either if the $\mathbf{\overline x}_i$ are $1$-sparse, in which case we may apply Thm.~\ref{DeterministicUniquenessTheorem} to guarantee both dictionary and code recovery, or if $\mathbf{B}$ has colinear columns. We can rule out the latter case because if that were so, we could consolidate all sets of colinear columns to form a dictionary with fewer than $m$ linearly independent columns and yet for which there must exist $1$-sparse $\mathbf{\overline x}_i$ solving Prob.~\ref{ConvexifiedOptimizationProblem}; but this would contradict the fact that $L_2(\mathbf{A}) > 0$. 
\end{proof}

%\begin{conjecture}\label{l1conjecture}
%Fix a matrix $\mathbf{A}$ with $\|\mathbf{Ae}_i\|_2 = 1$ for $i \in [m]$ and vectors $\mathbf{x}_i$ satisfying the assumptions of Thm.~\ref{SLCopt}. Every solution to Prob.~\ref{ConvexifiedOptimizationProblem} (with $\eta = \varepsilon/2$) and $\|\mathbf{Be}_i\|_2 = 1$. satisfies recovery guarantees \eqref{Cstable} and \eqref{b-PDa} when the corresponding bounds on $\eta$ are met.
%\end{theorem}

### 2020-08-13 manuscripts/thesis/introduction.tex
redid abstract

Formal statements of the main findings described above are given in Chap.~\ref{Results}, along with their adaptation to dictionaries and codes drawn from arbitrary (continuous) probability distributions (Thm.~\ref{robustPolythm} and Cor.~\ref{ProbabilisticCor}). All results assume real matrices and vectors. For clarity of exposition, the technical proofs of Thms.~\ref{DeterministicUniquenessTheorem} and ~\ref{SLCopt} are deferred to Chap.~\ref{DUT}, following some necessary definitions and the statement of a key lemma in combinatorial matrix analysis (Lem.~\ref{MainLemma}, proven in the chapter's Appendix). These results and their applications are discussed in Chap.~\ref{Discussion}, which concludes with some open questions and directions for future research seeded in part by practically-minded simulations. 
%The Appendix contains a proof of Lem.~\ref{MainLemma}.

>>>>

Formal statements of the main findings described above are given in Chap.~\ref{Results}, along with their adaptation to dictionaries and codes drawn from arbitrary (continuous) probability distributions (Thm.~\ref{robustPolythm} and Cor.~\ref{ProbabilisticCor}). All results assume real matrices and vectors. For clarity of exposition, the technical proofs of Thms.~\ref{DeterministicUniquenessTheorem} and ~\ref{SLCopt} are deferred to Chap.~\ref{DUT}, following some necessary definitions and the statement of a key lemma in combinatorial matrix analysis (Lem.~\ref{MainLemma}, proven in the chapter's Appendix). These results and their applications are discussed in Chap.~\ref{Discussion}, which concludes with some open questions and directions for future research seeded in part by some practically-minded simulations. 
%The Appendix contains a proof of Lem.~\ref{MainLemma}.

### 2020-08-13 manuscripts/thesis/results.tex
new intro
redid abstract

\subsection{Deterministic Uniqueness Theorems}

@@

\subsection{Probabilistic uniqueness theorems}

@@

As it seemed to benefit a reviewer of this work, some clarification may be in order on how the deterministic sample complexity $N = |\mathcal{H}\left[|(k-1){m \choose k} + 1\right]$ given here (Cor.~\ref{DeterministicUniquenessCorollary}) compares to those listed in the the top two rows listed in Table I of \cite{Hillar15}. To be clear, the theory developed here is strictly more general, since $\mathcal{H}$ can always be taken to be ${[m] \choose k}$. The point of difference in this comparison is the assumed set of supports for sparse codes, which is always ${[m] \choose k}$ in \cite{Hillar15}, whereas here it can be assumed to be any regular $k$-uniform hypergraph that satisfies the SIP. By row, %, assuming for our results that sufficient data have been sampled from all supports in a hypergraph H satisfying the SIP:
\begin{enumerate}
\item[I.] The result here improves upon the listed $k{m \choose k}^2$ by an exponential factor, since for every $k < m$ there exists a regular $k$-uniform hypergraph $\mathcal{H}$ with $|\mathcal{H}| = m$ satisfying the SIP. 
\item[II.] The authors in \cite{Hillar15} have applied measure-theoretic arguments to achieve $(k+1){m \choose k}$ with almost-certainty (i.e. with probability one), a factor of $m$ reduction over that for which certainty can alternatively be guaranteed here.    
%\item[III.] A direct comparison cannot be made because it is an open question as to the probability that a random set of supports satisfies the SIP. This would be an interesting problem for the community to solve; regardless, the result of \cite{Hillar15} is still implied by the more general theory given here.
%\item[IV.] The result here is $|\mathcal{H}\left[|(k-1){m \choose k} + 1\right]$ with probability one compared to $(k+1){m \choose k}$ with probability one. In this case the sample complexity stated in \cite{Hillar15} is marginally better, but it relies on a noise-free technicality unique to their assumption $\mathcal{H} = {m \choose k}$. %differs in flavor: their $\mathbf{x_i}$ are assumed to be distributed as $(k+1)$ samples per support for every support of size $k \in [m]$, whereas here it is assumed that $(k-1){m \choose k}$ codes $\mathbf{x_i}$ are distributed over each support in some hypergraph $\mathcal{H}$ satisfying the SIP. 
%\item[V.] Again, a direct comparison cannot be made for the reason stated for row III.
\end{enumerate}

>>>>

\subsection{Deterministic guarantees}

@@

\subsection{Probabilistic guarantees}

@@

As it seemed to benefit a reviewer of this work, some clarification may be in order on how the deterministic sample complexity $N = |\mathcal{H}|\left[(k-1){m \choose k} + 1\right]$ given here (Cor.~\ref{DeterministicUniquenessCorollary}) compares to those listed in the the top two rows listed in Table I of \cite{Hillar15}. To be clear, the theory developed here is strictly more general, since $\mathcal{H}$ can always be taken to be ${[m] \choose k}$. The point of difference in this comparison is the assumed set of supports for sparse codes, which is always ${[m] \choose k}$ in \cite{Hillar15}, whereas here it can be assumed to be any regular $k$-uniform hypergraph that satisfies the SIP. By row, %, assuming for our results that sufficient data have been sampled from all supports in a hypergraph H satisfying the SIP:
\begin{enumerate}
\item[I.] The result here improves upon the listed $k{m \choose k}^2$ by an exponential factor, since for every $k < m$ there exists a regular $k$-uniform hypergraph $\mathcal{H}$ with $|\mathcal{H}| = m$ satisfying the SIP. 
\item[II.] The authors in \cite{Hillar15} have applied measure-theoretic arguments to achieve $(k+1){m \choose k}$ with almost-certainty (i.e. with probability one), a factor of $m$ reduction over that for which certainty can alternatively be guaranteed here.    
%\item[III.] A direct comparison cannot be made because it is an open question as to the probability that a random set of supports satisfies the SIP. This would be an interesting problem for the community to solve; regardless, the result of \cite{Hillar15} is still implied by the more general theory given here.
%\item[IV.] The result here is $|\mathcal{H}\left[|(k-1){m \choose k} + 1\right]$ with probability one compared to $(k+1){m \choose k}$ with probability one. In this case the sample complexity stated in \cite{Hillar15} is marginally better, but it relies on a noise-free technicality unique to their assumption $\mathcal{H} = {m \choose k}$. %differs in flavor: their $\mathbf{x_i}$ are assumed to be distributed as $(k+1)$ samples per support for every support of size $k \in [m]$, whereas here it is assumed that $(k-1){m \choose k}$ codes $\mathbf{x_i}$ are distributed over each support in some hypergraph $\mathcal{H}$ satisfying the SIP. 
%\item[V.] Again, a direct comparison cannot be made for the reason stated for row III.
\end{enumerate}
